### Here is a sketch of design ideas for BUGS modules

This is implemented on branch `BUGSmodules`

```{r}
library(nimble)
nimbleOptions(enableBUGSmodules = TRUE) ## feature toggle
```

Here is the basic idea.  A declaration with BUGS module `lmPred` would get expanded in place:

```{r}
mc <- nimbleCode({
    a ~ dnorm(0,1)
    Ypred[1:n] <- lmPred(X[1:n])
})
```

expands to:

```{r}
nimble:::codeProcessBUGSmodules(mc)
```

This provides the linear predictors from `X[1:n]` to predict `Ypred[1:n]`.

When option `enableBUGSmodules == TRUE`, this will be done early in processing BUGS code:

```{r}
X <- 1:5
m <- nimbleModel(mc, constants = list(n = 5), data = list(X = X))
```

Note that the functionality in `lmPred` is currently a sham: It replaces the code but not by truly processing it.

One BUGS module can use another.  Say we want a linear model:

```{r}
mc <- nimbleCode({
    a ~ dnorm(0,1)
    Ypred[1:n] ~ nim_lm(X[1:n])
})
```

The non-recursed expansion uses the `lmPred` template:

```{r}
nimble:::nim_lm$process(Ypred[1:n], nim_lm(X[1:n])) ## The processing is fake, to show what the real result would look like.  It generates the response distributions and uses lmPred to expand linear predictions...
```

This means the recursed expansion has everything:

```{r}
nimble:::codeProcessBUGSmodules(mc)  ## ... which are then replaced by recursion
X <- 1:5
## The full expansion works in a model
m <- nimbleModel(mc, constants = list(n = 5), data = list(X = X))
```

It would be easy to remove extra `{`s and collapse identical `for` loops into one.

Similarly, a glm module could use `lmPred` but provide an argument to insert a link:

```{r}
mc <- nimbleCode({
    a ~ dnorm(0,1)
    Ypred[1:n] ~ nim_glm(X[1:n])
})
X <- 1:5
## non-recursed expansion, shows link argument
nimble:::nim_glm$process(Ypred[1:n], nim_glm(X[1:n])) ## In this case a link is requested in the linear expansion.
## recursed expansion shows full model
nimble:::codeProcessBUGSmodules(mc)  ## ... which is then replaced by recursion
## It works in a fully processed model
m <- nimbleModel(mc, constants = list(n = 5), data = list(X = X))
```

